### 2013-04-29 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding data pdf files

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Parameter and Estimator Analysis}
\label{sec:parameter_estimator_analysis}
With the adjoint method shown to be more effective than the direct
method and MCSA to have better iterative and timing performance than
sequential Monte Carlo, we now aim to study the effects of the adjoint
Monte Carlo parameters, weight cutoff and number of histories, on MCSA
performance with both the collision and expected value estimators.

@@

\subsection{Artificial Absorption}
\label{subsec:artificial_absorption}
The random walk sequences generated by the adjoint Neumann Ulam
decomposition in Eqs~(\ref{eq:adjoint_probability}) and
(\ref{eq:adjoint_weight}) will continue on indefinitely without the
implementation of a weight cutoff procedure (which in itself is
effectively a form of variance reduction). This is due to the fact
that all states to which a stochastic history may move in a given
transition exist within the system and have a non-zero weight. A
preliminary variance reduction scheme is to introduce
\textit{artifical absorption} into the system such that at each
transition event, a stochastic history will now have some finite
probability of being terminated through absorption as well as
transition to other states in the system. This probability, $p_{abs}$,
modifies the probabilities and weights in the adjoint random walk
sequence as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_i|h_{ji}|}(1-p_{abs})\:,
  \label{eq:absorption_probabilities}
\end{equation}
and
\begin{equation}
  w_{ij} = \frac{sign(h_{ji})\sum_i|h_{ji}|}{(1-p_{abs})}\:.
  \label{eq:absorption_weights}
\end{equation}
At each transition step, a history at state $i$ will then sample the
probability distribution function generated by
Eq~(\ref{eq:absorption_probabilities}) with an additional absorption
state added to the distribution. If the sampling procedure results in
a transition to the absorption state, the history is immediately
terminated. The advantages of this for residual Monte Carlo are that
increasing the absorption probability decreases the length of each
random walk (resulting in speedup), while the random walk has a higher
weight contribution at every step. This higher weight means that each
history is depositing more information near its birth site which will
on average be located where the system residual is the largest. An
additional advantage of this formulation is that the full Neumann Ulam
decomposition is maintained, resulting in Monte Carlo estimates that
are unbiased.

To observe the effects of this variance reduction on MCSA solutions,
the transient Poisson problem was again solved with the adjoint
method, this time with varying values of artificial
absorption. Compare CPU time and iterative performance with a fixed
grid.

>>>>

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Parameter and Estimator Analysis}
\label{sec:parameter_estimator_analysis}
With the adjoint method shown to be more effective than the direct
method and MCSA to have better iterative and timing performance than
sequential Monte Carlo, we now aim to study the effects of the adjoint
Monte Carlo parameters, weight cutoff and number of histories, on MCSA
performance with both the collision and expected value
estimators. With the same Poisson problem, we will use a $200 \times
200$ grid and a convergence tolerance of \sn{1}{-8} for all
calculations. To study the effects of the number of Monte Carlo
histories per MCSA iteration, the weight cutoff was fixed at
\sn{1}{-4} while the number of histories per iteration was varied from
6,000 to 100,000. It was observed that MCSA would not converge for
this problem using the collision estimator with less than 6,000
histories while the expected value estimator permitted convergence
with only 2,000 histories. Figure~\ref{fig:estimator_nh_iters} gives
the number of iterations required to converge for both estimators as a
function of the number of histories per iteration. For smaller numbers
of histories, the performance of the expected value estimator is
significantly better than the collision estimator. This result is
valuable in that less transport is required to achieve the same MCSA
iterative performance with the expected value estimator, important for
situations where transport is expensive (i.e. in domain decomposed
calculations). Interstingly, as the number of histories per iteration
are increased, the iterative performance with the collision estimator
approaches that of the expected value estimator. However, given the
performance of the expected value estimator at a fractional number of
histories, one should strongly consider its use over using the collision
estimator with more histories. The CPU time required to converge is
presented in Figure~\ref{fig:estimator_nh_time} and reflects the
results of the iterative performance. In general, using the collision
estimator is slightly slower overall, but the time per iteration is
faster due to the fact that the estimator does not have to cycle
through multiple states during the tally procedure.

\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. Monte Carlo
      histories per MCSA iteration for a $200 \times 200$ square mesh
      and a weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
      histories, the expected value estimator performance is
      significantly better than the collision estimator. At higher
      numbers of histories, the estimators become roughly
      equivalent.}}
  \label{fig:estimator_nh_iters}
\end{figure}

\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_nh_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Monte Carlo histories
      per MCSA iteration for a $200 \times 200$ square mesh and a
      weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
      histories, the expected value estimator performance is better
      than the collision estimator due to a lower iteration count
      while the actual compute time per iteration is higher.}}
  \label{fig:estimator_nh_time}
\end{figure}

For the weight cutoff study, the number of histories per iteration was
fixed at 40,000 and the weight cutoff varied from \sn{5}{-1} down to
\sn{1}{-10}. Surprisingly, the number of iterations to converge given
by Figure~\ref{fig:estimator_wc_iters} is effectively invariant to the
weight cutoff, only seeing detrimental effects on the number of
iterations at a very large weight cutoff. This suggests that a fairly
large weight cutoff can be used in practice with both estimators and
that the preliminary components of the random walk are more important
to MCSA convergence than those that occur later and with lower weight
contributions. To further motivate using a larger weight cutoff,
Figure~\ref{fig:estimator_wc_time} gives the CPU time need to converge
as a function of weight cutoff. As expected, lowering the weight
cutoff lengthens the random walk lengths and increases CPU time at no
gain of iterative performance. In general, these results suggest using
the expected value estimator over the collision estimator with MCSA as
well as using a larger weight cutoff.

\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. history weight
      cutoff for a $200 \times 200$ square mesh and 40,000 histories.}
    \textit{For low numbers of histories, the expected value estimator
      performance is significantly better than the collision
      estimator. At higher numbers of histories, the estimators become
      roughly equivalent.}}
  \label{fig:estimator_wc_iters}
\end{figure}

\begin{figure}[p!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_wc_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. history weight cutoff
      for a $200 \times 200$ square mesh and 40,000 histories.}
    \textit{For low numbers of histories, the expected value estimator
      performance is better than the collision estimator due to a
      lower iteration count while the actual compute time per
      iteration is higher.}}
  \label{fig:estimator_wc_time}
\end{figure}

@@

\subsection{Artificial Absorption}
\label{subsec:artificial_absorption}
The random walk sequences generated by the adjoint Neumann Ulam
decomposition in Eqs~(\ref{eq:adjoint_probability}) and
(\ref{eq:adjoint_weight}) will continue on indefinitely without the
implementation of a weight cutoff procedure (which in itself is
effectively a form of biased variance reduction). This is due to the
fact that all states to which a stochastic history may move in a given
transition exist within the system and have a non-zero weight. A
preliminary variance reduction scheme is to introduce
\textit{artifical absorption} into the system such that at each
transition event, a stochastic history will now have some finite
probability of being terminated through absorption as well as
transition to other states in the system. This probability, $p_{abs}$,
modifies the probabilities and weights in the adjoint random walk
sequence as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j|h_{ji}|}(1-p_{abs})\:,
  \label{eq:absorption_probabilities}
\end{equation}
and
\begin{equation}
  w_{ij} = \frac{sign(h_{ji})\sum_j|h_{ji}|}{(1-p_{abs})}\:.
  \label{eq:absorption_weights}
\end{equation}
At each transition step, a history at state $i$ will then sample the
probability distribution function generated by
Eq~(\ref{eq:absorption_probabilities}) with an additional absorption
state added to the distribution. If the sampling procedure results in
a transition to the absorption state, the history is immediately
terminated. The advantages of this for residual Monte Carlo are that
increasing the absorption probability decreases the length of each
random walk (resulting in speedup), while the random walk has a higher
weight contribution at every step. This higher weight means that each
history is depositing more information near its birth site which will
on average be located where the system residual is the largest. An
additional advantage of this formulation is that the full Neumann Ulam
decomposition is maintained, resulting in Monte Carlo estimates that
are unbiased.

To observe the effects of this variance reduction on MCSA solutions,
the transient Poisson problem was again solved with the adjoint
method, this time with varying values of artificial absorption. A $200
\times 200$ grid was used with 40,000 histories at every iterations
with the expected value estimator. To reduce the effect of the weight
cutoff on this study, the weight cutoff was set to \sn{1}{-12} such
that it is signicantly more likely that a history will be terminated
by absorption rather than weight cutoff. Compare CPU time and
iterative performance with a fixed grid.
